LOSD: A Linked Open Social Dataset for Scientific Benchmarking

### 2019-07-01 mdpi/template.tex
arrumados trabalhos relacionados e bibliografia

\begin{document}
\section{Summary}
The research on human social and complex networks often
rely on data that express real phenomena~\cite{c1,c2,c3}.
Such data is regularly not made publicly available,
which hinders validation of the results
and does not favor subsequent work.
Furthermore, there is a lack of open datasets for benchmarking results
related to complex and social networks,
yielding diverse results from poorly related sources.
Social networks linked data is currently scarce and not stable
or rigorous enough to be a public reference on statistical physics
and social networks research~\cite{l1,l2}.

This work presents a linked and open dataset
derived from online human social networks.
The dataset is named LOSD (Linked Open Social Data),
and has been achieved by linking social data into a dataset derived from diverse
provenance, namely: Facebook, Twitter, IRC, Email groups, ParticipaBR, Cidade Democrática, and AA.
Such data may be of use as a common and public repertoire for scientific
research involving complex networks, specially if social networks and textual content are pinpointed.
In fact, the data has already been used in scientific research~\cite{stab,thesis} (CNPq grant ),
and should be further used e.g. in researching complex networks and data visualization (FAPESP grant )
which relies on networks enriched by textual data and on dynamic (or time-evolving) networks.

Data collection was performed by a potentially innovative method which places the researcher
at the core of data collection, and keeps the resources (e.g. code, data, documents) open,
in order to ammeliorate ethical issues involved in studying human social systems.

This paper is organized as follows. Next subsection briefly discusses related work.
Section~\ref{materials} describes the data currently available.
Section~\ref{methods} describes the methods used for data gathering, translation into RDF linked data, and the synthesis of auxiliary structures.
Section~\ref{usage} contains direction for using the data.

\subsection{Related work}
Most available datasets on human social networks rely on only one source~\cite{o1,o2,o3,o4},
on specialized communities such as related to research or movies~\cite{s1,s2,s3,s4,s5},
or present only statistics about social networks, and not the data that yields the networks~\cite{st1,st2,st3,st4}.
Contributions which are most similar to LOSD are mentioned in this section.
In~\cite{fb1}, data obtained only from Facebook is reported.
With such a focus, the authors made available a number of statistics,
but the data is offline at least since 2010.
Linked human social network data is scarcely found in the scientific community.
In~\cite{foaf1}, the authors presented a set of tools to obtain linked social data available in the public data cloud, and made available a dataset from the Advogato platform,
dedicated to free software developers linked by trust relatioships.
There are various datasets which makes available data obtained from Twitter (e.g.~\cite{tw1,tw2,t3,t4}),
most often with a specific focus (users, retweets, hashtags, etc)
and not presented to the scientific community by means of a research article.
In summary, no human social network dataset presented to the
scientific community containts the diverse provenance used in LOSD,
potentially because researchers did not have the means to acquire and
share the data as described in Section~\ref{acq}.

\section{Data Description}\label{materials}
\subsection{Data outline}\label{outline}
% scripts/percolationUse2.py
The database consists of 31,996,740 triples, 3,030,433 network links yield by main interactions or relations, 350,284 participants and 245,194,377 characters. Among all snapshots, 63 are ego snapshots, 54 are group snapshots; 50 have interaction edges, 89 have friendship edges; 43 have text content from messages.
Appendix~\ref{aoverall} holds the queries used to obtain such data.
This section describes de data in LOSD in very brief terms.
The Supplementary document is provided for further information about the dataset.

@@

\subsection{Data acquisition}\label{acq}
Research involving human subjects raise well-known ethic issues~\cite{eth}.
Furthermore, the online social networking platforms often restrict access
to the data and declare illegal the use of their data through web crowlers~\cite{ile1,ile2}.
Although Twitter and GMANE (email) data is available through public APIs,
the other sources are more problematic.
In this context, research using data, e.g. from Facebook, is precarious
if not prohibited.
We reached a potential solution to this issue through continuated research with
collaborators in the fields of phylosophy, social sciences, antropology and psychology.
The conceptual framework is presented in~\cite{antphy,antphy2}, and in essence
consists in researching first the social networks related to the researchers,
and keeping the procedures, data, algorithms and results publicly available.
These practices enabled collecting and sharing the data in LOSD,
and was called ``Anthropological Physics'' because it relies in the
etnographic tecnique of investigating the social structures
by studying the researcher, as to ammeliorate the ethic issues.
Accordingly, aside from the data acquired from Twitter, all data in LOSD
was gathered from social instances where the first author is a participant.
For example, all Facebook ego networks were donated to the first author by
users which know him and were interested in his research and intervention on the networks,
and all email and Facebook groups data was obtained from groups where the first author is a member.

@@

\subsection{RDF}
The Resource Description Framework (RDF), a W3C
recommendation, is a model for data
interchange~\cite{w3cr}.
It is based on the idea of making statements about resources in the form
of triples, i.e. expressions in the form ``subject - predicate -
object''.
RDF can be serialized in several file formats, including RDF/XML,
Turtle and Manchester, which all, in essence, represent a labeled and
directed multi-graph.
RDF may be stored in a type of database called a triplestore~\cite{rdf}.

@@

One can also write an ontology using the OWL protocol,
with which all the expressive capabilities of RDFS are
available, but one can also, among other things~\cite{owl}:
\begin{itemize}
  \item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
  \item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily holds a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.
Using ontologies enables automated reasoning.
For example, if a property \textttt{:r} is known to relate only mokeys to trees,
then, if there is a triple \textttt{:a :r :b}, the instance \textttt{:a} is considered a monkey,
and instance \textttt{:b} is considered a tree,
even if there is no explicit declaration similar
to \textttt{:a rdf:type :Mokey}.

For the data described in this article, the following features are most relevant~\cite{rdfs,owl}:
\begin{itemize}
  \item Using RDFS, one may define a property domain, i.e. the class instances that are allowed as the object in a triple with the property.
  \item Using RDFS, one may define a property range, i.e. the class instances that are allowed as the predicate in a triple with the property.
  \item Using OWL, one may define an existential restriction for a class \textttt{:C},
    i.e. declare that an instance \textttt{:C\#I} of such class has at least one relation with a property \textttt{:p} and a class \textttt{:C2} in the form
    \textttt{:C\#I :p :C2\#I}, where \textttt{:C2\#I} is an instance of the class \textttt{:C2}.
  \item Using OWL, one may define an universal restriction for a class \textttt{:C},
    i.e. declare that any relation of an instance \textttt{:C\#I} of such class, with a property \textttt{:p} of the form \textttt{:C\#I :p :CX\#I},
    implies that \textttt{:CX\#I} is an instance of a specific class, e.g. \textttt{:C2}.
\end{itemize}

\subsection{Data-driven ontology synthesis}\label{ont}
Most ontologies are created by domain experts even though the data they
arrange is often given by a software system and has a predefined
structure~\cite{owl1,owl2}.  
We developed a simple ontology synthesis method that probes
the ontological structure in the data with
SPARQL queries and post-processing.
The results are OWL code and diagrams which are available in the
Supplementary document of this article.
The method can be extended to comprise further OWL axioms and restrictions,
but is currently performed to fit present needs with maximum simplicity.
Present needs are limited to informative figures and
the steps implemented are as follows:
\begin{enumerate}[leftmargin=0cm]
    \item Obtain all distinct classes with the query:\\
        \h\textttt{SELECT DISTINCT ?class\_uri WHERE \{ ?s a ?class\_uri .\}}
\item For each class, obtain the properties that occur as predicates in triples where the subject is an instance of the class:\\
	\h\textttt{SELECT DISTINCT ?property\_uri WHERE \{ ?s a <class\_uri> . ?s ?property\_uri ?o .\}}\\
Such properties are used to assert existential and universal restrictions for the class.
\item Compare the total number of individuals (\textttt{?cs1}) of the class (\textttt{class\_uri}) with
	the number of such individuals (\textttt{?cs2}) that are subjects of at least one triple where 
        the predicate is the property (\textttt{property\_uri}) and the object is
        an instance of the same class (\textttt{class2\_uri}).
	The queries are:\\
	\h\textttt{SELECT (COUNT(DISTINCT ?s) as ?cs1) WHERE \{ ?s a <class\_uri> \}}\\
        \h\textttt{SELECT DISTINCT ?oc WHERE \{\\
	\h\h ?s a <class\_uri>. ?s <property\_uri> ?o . ?o a ?oc . \\ 
        \h\}}\\
        Then, for each object class (\textttt{object\_curi}) in \textttt{?oc}:\\
	\h\textttt{SELECT (COUNT(DISTINCT ?s) as ?cs2) WHERE \{\\
	\h\h ?s a <class\_uri>. ?s <property\_uri> ?o . ?o a <object\_curi> .\\         \h\}}\\
        When \textttt{?cs1} matches \textttt{?cs2}, an existential restriction is found. And with the query:
	\h\textttt{SELECT (COUNT(DISTINCT ?s) as ?cs) WHERE \{\\
	\h\h ?s a <class\_uri> . ?s <property\_uri> ?o . ?o a ?oc .\\
        \h\h FILTER(str(?oc) != 'object\_curi')\\
        \h\}}\\
       When \textttt{?cs} is 0, a universal restriction is found. 
\item To keep a record of further restrictions, get all object classes or datatypes where the subject is an instance of the class and the predicate is the property:\\
	\h\textttt{SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE \{\\
        \h\h ?s a <class\_uri>. ?s <property\_uri> ?o . OPTIONAL \{ ?o a ?co . \}\\
        \h\}}
    \item Obtain all distinct properties:\\
        \h\textttt{SELECT DISTINCT ?p WHERE \{ ?s ?p ?o \}}
    \item Check if each property is functional, i.e. if it
        occurs at most once with each subject.
        This is performed by counting the objects and further verifying
        that they are at most one. The query is:\\
        \h\textttt{SELECT DISTINCT (COUNT(?o) as ?co) WHERE \{ ?s
            <property\_uri> ?o \} GROUP BY ?s}
          \item For each property, find the incident range (possible subjects) and domain (possible objects) with the
        queries:\\
        \h\textttt{SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE \{\\
        \h\h ?s <property\_uri> ?o . OPTIONAL \{ ?o a ?co . \}\\
        \h\}} \\
        and \\
        \h\textttt{SELECT DISTINCT ?cs WHERE \{ ?s <property\_uri> ?o . ?s a ?cs . \}}
%     \item For each instance of each class, get all distinct predicates.
%         For each predicate, check if all instances of the class
%         hold such relationship (existential restriction):\\
%         \textttt{SELECT DISTINCT ?p WHERE \{ ?s a <class\_uri>. ?s ?p ?o . \}}\\
%         \textttt{SELECT DISTINCT ?s WHERE \{ ?s a <class\_uri> \}}\\
%         \textttt{SELECT DISTINCT ?s ?co  (datatype(?o) as ?do) WHERE \{?s
%                 a <class\_uri>. ?s <property\_uri> ?o . OPTIONAL \{?o a ?co . \}\}}
%     \item and if all instances that hold such relationship are instances of the class
%         (universal restriction):\\
%         \textttt{SELECT DISTINCT ?s WHERE \{ ?s <property\_uri> ?o . \}}
%     \item Draw each class, each property and the overall figure.
      \item Render diagrams to achieve figures as exemplified in Section~\ref{sdia} and in the Supporting Information document.
%      \item Make \textttt{rdfs:subClassOf} and \texttt{rdfs:subPropertyOf}
%          statements to better organize knowledge and link to third party
%          ontologies and data. This step is left aside in current LOSD as 
\end{enumerate}

\subsection{Standardization}
The data is embedded into URIs and triples, i.e. translated to RDF.
URIs are built in the namespace \url{http://purl.org/socialparticipation/participationontology/}
which are identified herein with the prefix ``\textttt{po:}''.
Classes and properties are built by adding a suffix to the root, as in \textttt{po:Participant} or \textttt{po:text}.
Classes have ``UpperCamelCase'' suffixes while properties have ``lowerCamelCase'' suffixes.
All class instances, such as participants, messages, friendships and
interactions, are linked to
snapshots through the triple \textttt{<instance> po:snapshot <snapshot\_uri>}.
Message texts, including comments, are objects in the triple: \textttt{<message\_id> po:text <message\_text>}.
Preprocessed texts are objects of triples: \textttt{<message\_id> po:cleanText <message\_text>}.
More specialized predicates are used for delivering text when necessary,
such as \textttt{po:htmlBodyText} and \textttt{po:cleanBodyText} used
for ParticipaBR articles.
A participant URI is unique throughout the provenance (e.g. the same for
the same participant in all Twitter snapshots).
To enable annotations which differ when the snapshot changes,
\texttt{po:Observation} class instances are used in the triple
\textttt{<participant\_uri> po:observation <observation\_uri>}.
The observation instances are then linked to the snapshot and the
data.

Instance URIs are built using the URI from the class they derive from plus a hashtag character,
a provenance string (e.g. \textttt{facebook-legacy} or
\textttt{participabr-legacy}), and an identifier;
i.e. \textttt{po:Participant\#<provenance-legacy>-<id>}.
All snapshot URIs follow the formation rule: \textttt{po:<SnapshotProvenance>\#<snapshot\_id>}.
All snapshot ids follow the formation rule: \textttt{<platform>-legacy-<further\_identifier>}; e.g.
\textttt{irc-legacy-labmacambira} or
\textttt{email-legacy-linux.audio.devel1-20000}.

\section{Usage notes}\label{usage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt} 
\subsection{Software tools}
The authors made available software for rendering analyses and
the data set itself.
\subsubsection{Triplification routines}
For each social platform there is a \emph{triplification} routine,
i.e. a script for expressinf the data as RDF.
Original formats and further observations are presented in
Table~\ref{tab:provenance}.
  {\renewcommand{\arraystretch}{1.2}
\begin{table*}[h!]\scriptsize
\begin{center}
\caption{Social platforms, original formats and further observations for
the database.}\label{tab:provenance}
\begin{tabular}{ l | p{3cm} p{3cm} c }
    \textbf{social platform} & \textbf{original format} & \textbf{further observations} & \textbf{toolbox} \\\specialrule{1.5pt}{1pt}{1pt}
    AA & MySQL and MongoDB databases; IRC text logs & donated by AA users & Participation~\cite{participation} \\\hline
    Cidade Democrática & MySQL database & donated by admins & Participation \\\hline
    Email & mbox & obtained through the Gmane public archive & Gmane~\cite{gmane} \\\hline
    Facebook & GDF, GML and TAB & obtained through Netvizz~\cite{netvizz} & Social~\cite{social} \\\hline
    IRC & plain text log & obtained through Supybot logging & Social \\\hline
    ParticipaBR & PostgreSQL database & donated by admins & Participation \\\hline
    Twitter & JSON & obtained through Twitter streaming API & Social \\
\end{tabular}\end{center}
\end{table*}                    
  }
\subsubsection{Topological and textual analysis}\label{ana}
Routines are available for taking topological and textual measures from
the LOSD data.
  Auxiliary routines, such as for performing principal component analysis (PCA)
  and obtaining statistics derived from the Kolmogorov-Smirnov test~\cite{kol}, are available to ease pattern recognition.

% \subsubsection{Multimedia rendering}\label{media}
% It is a core purpose of framework to provide routines for rendering
% audiovisualizations of the data.
% Social structures are rendered into music, images and video animations
% through the Percolation toolbox~\cite{percolation} in association with
% the Music and Visuals toolboxes~\cite{music,visuals}.

% \subsubsection{Migration from deprecated toolboxes}
% Routines mentioned in Sections~\ref{ana} and~\ref{media} are being migrated from deprecated
% toolboxes~\cite{gmaneLegacy,percolationLegacy} into newly designed
% toolboxes~\cite{percolation,visuals}. TTM, recover?

\subsection{Diagrams of the data and auxiliary tables}\label{sdia}
LOSD navigation may be assisted through diagrams which describes
the structure from each provenance.
Such diagrams are in the Supporting Information document
together with tables to facilitate the appreciation of the provided data.
A very simple example is given in Figure~\ref{dia} where the friendship
structure of the Facebook snapshots is revealed.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=0.5\textwidth]{../ontologies/facebook-legacy-AntonioAnzoategui18022013Friendship.ttl/draw}
    \caption{A diagram of the core structure that entails the friendship networks
    of the Facebook snapshots. A green edge denotes an OWL existential class restriction; an inverted nip denotes an OWL universal class restriction; a full (non-dashed) edge denotes an OWL functional property axiom (see Section~\ref{sont}). Further information and complete diagrams for each provenance are available in the Supporting Information document.}\label{dia}
\end{figure}

>>>>

\begin{document}
\section{Summary}
The research on human social and complex networks often
rely on data that express real phenomena~\cite{c1,c2,c3}.
Such data is regularly not made publicly available,
which hinders validation of the results
and does not favor subsequent work.
Furthermore, there is a lack of open datasets for benchmarking results
related to complex and social networks,
yielding diverse results from poorly related sources.
Social networks linked datasets are currently most often restricted to a specific platform or express statistics of social networks, as described in Section~\ref{srel}.
This work presents 
a dataset named LOSD (Linked Open Social Data)
that has been achieved by linking social data derived from diverse
provenance, namely: Facebook, Twitter, IRC, Email groups, ParticipaBR, Cidade Democrática, and AA.
Such data may be of use as a common and public repertoire for scientific
research involving complex networks, specially if social networks and textual content are pinpointed.
In fact, the data has already been used in scientific research~\cite{stab,thesis} (CNPq grant 140860/2013-4),
and should be further used e.g. in researching complex networks and data visualization (FAPESP grant 2017/05838-3)
which relies on networks enriched by textual data and on dynamic (or time-evolving) networks.
Data collection was performed by a potentially innovative method which places the researcher
at the core of data collection, and keeps the resources (e.g. code, data, documents) open,
in order to ammeliorate ethical issues involved in studying human social systems.

This paper is organized as follows. Next subsection briefly discusses related work.
Section~\ref{materials} describes the data currently available.
Section~\ref{methods} describes the methods used for data gathering, translation into RDF linked data, and the synthesis of auxiliary structures.
Section~\ref{usage} holds directions for using the data.

\subsection{Related work}\label{srel}
Most available datasets on human social networks rely only on one source (e.g.~\cite{nat,fb1}),
on specialized communities such as related to research or movies (e.g.~\cite{s1},
or present only statistics about social networks, and not the data that yields the networks (e.g.~\cite{st1}.
Linked human social network data is scarcely found in the scientific community.
In~\cite{foaf1}, the authors presented a set of tools to obtain linked social data available in the public data cloud, and made available a dataset from the Advogato platform,
dedicated to free software developers linked by trust relatioships.
There are various datasets which makes available data obtained from Twitter (e.g.~\cite{tw1,tw2}),
most often with a specific focus (users, retweets, hashtags, etc)
and not presented to the scientific community by means of a research article.
The datasets which are most closely related to LOSD are KONECT~\cite{konect},
SNAP Datasets~\cite{snapnets}, the Network Repository~\cite{nr}, and ICON~\cite{icon}.
None of them are (RDF) linked data but express the networks in formats specialized for networks.
In total, these datasets make available thousands of social networks, from diverse provenance and of various types (e.g. simple, directed, weighted, bipartite, temporal networks).
Compared to LOSD, KONECT, SNAP, Network Repository, and ICON are complementary, providing a further diversity of networks, but only as simplified data, i.e. in specialized formats, most often expressed as columns of node ids that are connected and metadata (such as timestaps for temporal networks).
In summary, no linked human social network dataset presented to the
scientific community containts the diverse provenance used in LOSD,
or provides such a wide range of metadata,
potentially because researchers did not have the means to acquire and
share the data as described in Section~\ref{acq},
and because linked data standards are (still) not widely adopted in 
every research domain.

\section{Data Description}\label{materials}
\subsection{Data outline}\label{outline}
% scripts/percolationUse2.py
The database consists of 31,996,740 triples, 3,030,433 network links yield by main interactions or relations, 350,284 participants and 245,194,377 characters. Among all snapshots, 63 are ego snapshots, 54 are group snapshots; 50 have interaction edges, 89 have friendship edges; 43 have text content from messages.
Section~\ref{queries} holds the queries used to obtain such data.
This section describes de data in LOSD in very brief terms.
The Supplementary document is provided for further information about the dataset.

@@

\subsection{Data acquisition}\label{acq}
Research involving human subjects raise well-known ethical issues,
although ethical regulation for (non-medical) research in social systems is not always endorsed by scientists~\cite{eth}.
Furthermore, the online social networking platforms often restrict access
to the data and user may consider illegal the use of their data through web crowlers~\cite{ile1,ile2,ile3}.
Although Twitter and GMANE (email) data is available through public APIs,
the other sources are more problematic.
In this context, research using data, e.g. from Facebook, is precarious
if not prohibited.
We reached a potential solution to this issue through continuated research with
collaborators in the fields of phylosophy, social sciences, antropology and psychology.
The conceptual framework is presented in~\cite{antphy,antphy2}, and in essence
consists in researching first the social networks related to the researchers,
and keeping the procedures, data, algorithms and results publicly available.
These practices enabled collecting and sharing the data in LOSD,
and was called ``Anthropological Physics'' because it relies in the
etnographic tecnique of investigating the social structures
by studying the researcher, as to ammeliorate the ethic issues.
Accordingly, aside from the data acquired from Twitter, all data in LOSD
was gathered from social instances where the first author is a participant.
For example, all Facebook ego networks were donated to the first author by
users which know him and were interested in his research and intervention on the networks,
and all email and Facebook groups data was obtained from groups where the first author is a member.

@@

\subsection{RDF}
The Resource Description Framework (RDF), a W3C
recommendation, is a model for data
interchange.
It is based on the idea of making statements about resources in the form
of triples, i.e. expressions in the form ``subject - predicate -
object''.
RDF can be serialized in several file formats, including RDF/XML,
Turtle and Manchester, which all, in essence, represent a labeled and
directed multi-graph.
RDF may be stored in a type of database called a triplestore.~\cite{rdf}

@@

One can also write an ontology using the OWL protocol,
with which all the expressive capabilities of RDFS are
available, but one can also, among other things~\cite{ont}:
\begin{itemize}
  \item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
  \item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily holds a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.
Using ontologies enables automated reasoning.
For example, if a property \textttt{:r} is known to relate only mokeys to trees,
then, if there is a triple \textttt{:a :r :b}, the instance \textttt{:a} is considered a monkey,
and instance \textttt{:b} is considered a tree,
even if there is no explicit declaration similar
to \textttt{:a rdf:type :Mokey}.

For the data described in this article, the following features are most relevant~\cite{rdfs,ont}:
\begin{itemize}
  \item Using RDFS, one may define a property domain, i.e. the class instances that are allowed as the object in a triple with the property.
  \item Using RDFS, one may define a property range, i.e. the class instances that are allowed as the predicate in a triple with the property.
  \item Using OWL, one may define an existential restriction for a class \textttt{:C},
    i.e. declare that an instance \textttt{:C\#I} of such class has at least one relation with a property \textttt{:p} and a class \textttt{:C2} in the form
    \textttt{:C\#I :p :C2\#I}, where \textttt{:C2\#I} is an instance of the class \textttt{:C2}.
  \item Using OWL, one may define an universal restriction for a class \textttt{:C},
    i.e. declare that any relation of an instance \textttt{:C\#I} of such class, with a property \textttt{:p} of the form \textttt{:C\#I :p :CX\#I},
    implies that \textttt{:CX\#I} is an instance of a specific class, e.g. \textttt{:C2}.
\end{itemize}

\subsection{Data-driven ontology synthesis}\label{ont}
Most ontologies are created by domain experts~\cite{ont}
even though the data they arrange is often given by a software system
and has thus a predefined structure.  
We developed a simple ontology synthesis method that probes
the ontological structure in the data with
SPARQL queries and post-processing.
The results are OWL code and diagrams which are available in the
Supplementary document of this article.
The method can be extended to comprise further OWL axioms and restrictions,
but is currently performed to fit present needs with maximum simplicity.
Present needs are limited to informative figures and
the steps implemented are as follows:
\begin{enumerate}
    \item Obtain all distinct classes with the query:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?class_uri WHERE { ?s a ?class_uri .}
\end{lstlisting}
\item For each class, obtain the properties that occur as predicates in triples where the subject is an instance of the class:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?property_uri WHERE { ?s a <class_uri> . ?s ?property_uri ?o .}
\end{lstlisting}
\item Compare the total number of individuals (\te{?cs1}) of the class (\te{class\_uri}) with
	the number of such individuals (\te{?cs2}) that are subjects of at least one triple where 
        the predicate is the property (\te{property\_uri}) and the object is
        an instance of the same class (\te{class2\_uri}).
	The queries are:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs1) WHERE { ?s a <class_uri> }
SELECT DISTINCT ?oc WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
}
\end{lstlisting}
        Then, for each object class (\textttt{object\_curi}) in \textttt{?oc}:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs2) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a <object_curi> .
}
\end{lstlisting}
        When \te{?cs1} matches \te{?cs2}, an existential restriction is found.
        And with the following query, when \te{?cs} is 0, a universal restriction is found:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?s) as ?cs) WHERE {
  ?s a <class_uri> . ?s <property_uri> ?o . ?o a ?oc .
  FILTER(str(?oc) != 'object_curi')
}
\end{lstlisting}
  \item Retrieve all object classes or datatypes where the subject is an instance of the class and the predicate is the property:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s a <class_uri>. ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
}
\end{lstlisting}
  \item Obtain all distinct properties:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?p WHERE { ?s ?p ?o }
\end{lstlisting}
  \item For each property, check if it is functional, i.e. if it
        occurs at most once with each subject.
        This is performed by counting the objects and further verifying
        that they are at most one. The query is:
\begin{lstlisting}[language=spq]
SELECT DISTINCT (COUNT(?o) as ?co) WHERE {
  ?s <property_uri> ?o 
} GROUP BY ?s
\end{lstlisting}
  \item For each property, find the incident range (possible subjects) and domain (possible objects) with the
        queries:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?co (datatype(?o) as ?do) WHERE {
  ?s <property_uri> ?o . OPTIONAL { ?o a ?co . }
} 
SELECT DISTINCT ?cs WHERE { ?s <property_uri> ?o . ?s a ?cs . }
\end{lstlisting}
  \item Render diagrams to achieve figures as exemplified in Section~\ref{sdia} and in the Supporting Information document.
\end{enumerate}

\subsection{Standardization}
LOSD data is represented through URIs and triples by translation of various formats to RDF.
URIs are built in the namespace \url{http://purl.org/socialparticipation/participationontology/}
which are identified herein with the prefix ``\te{po:}''.
Classes and properties are built by adding a suffix to the root, as in \te{po:Participant} or \te{po:text}.
Classes have ``UpperCamelCase'' suffixes while properties have ``lowerCamelCase'' suffixes.
All class instances, such as participants, messages, friendships and
interactions, are linked to
snapshots through the triple \te{<instance> po:snapshot <snapshot\_uri>}.
Message texts, including comments, are objects in a triple: \te{<message\_id> po:text '...the text...'}.
Preprocessed texts are objects of triples: \te{<message\_id> po:cleanText '...te text...'}.
More specialized predicates are used for text storage when necessary,
such as \te{po:htmlBodyText} and \te{po:cleanBodyText} used
for ParticipaBR articles.
A participant URI is unique throughout the provenance (e.g. the same for
the same participant in all Twitter snapshots).
To enable annotations which differ when the snapshot changes,
\te{po:Observation} class instances are used in the triple
\te{<participant\_uri> po:observation <observation\_uri>}.
Observation instances are then linked to the snapshot and the
data.

Instance URIs are built using the URI from the class they derive from plus a hashtag character,
a provenance string (e.g. \te{facebook-legacy} or
\textttt{participabr-legacy}), and an identifier;
e.g. \textttt{po:Participant\#<provenance-legacy>-<id>}.
All snapshot URIs follow the formation rule: \textttt{po:<SnapshotProvenance>\#<snapshot\_id>}.
All snapshot ids follow the formation rule: \textttt{<platform>-legacy-<further\_identifier>}; e.g.
\textttt{irc-legacy-labmacambira} or
\textttt{email-legacy-linux.audio.devel1-20000}.

The authors chose to express all URIs in LOSD locally (i.e. within a local namespace), and linkage with external namespaces may be achieved e.g. by using subclass and ``same as'' relations.
Also, each snapshot holds specific \te{po:observation} for each participant, allowing for specific metadata such as provided by the original data or as may be provided uppon analysis.
LOSD does not hold data about each snapshot (such as statistics of the networks), and linkage to external namespaces (such as FOAF and Dublin Core)
to avoid overcomplicating this initial contribution and unecessary use of storage space.

\section{Usage notes}\label{usage}

\subsection{Triplification routines}
For each social platform there is a \emph{triplification} routine,
i.e. a script for expressing the data as RDF.
Original formats and further information are presented in
Table~\ref{tab:provenance}.
  {\renewcommand{\arraystretch}{1.2}
\begin{table*}[h!]\scriptsize
\begin{center}
\caption{Social platforms, original formats and further observations for
the database.}\label{tab:provenance}
\begin{tabular}{ l | p{3cm} p{3cm} c }
    \textbf{social platform} & \textbf{original format} & \textbf{further observations} & \textbf{toolbox} \\\specialrule{1.5pt}{1pt}{1pt}
    AA & MySQL and MongoDB databases; IRC text logs & donated by AA users & Participation~\cite{participation} \\\hline
    Cidade Democrática & MySQL database & donated by admins & Participation \\\hline
    Email & mbox & obtained through the Gmane public archive & Gmane~\cite{gmane} \\\hline
    Facebook & GDF, GML and TAB & obtained through Netvizz~\cite{netvizz} & Social~\cite{social} \\\hline
    IRC & plain text log & obtained through Supybot logging & Social \\\hline
    ParticipaBR & PostgreSQL database & donated by admins & Participation \\\hline
    Twitter & JSON & obtained through Twitter streaming API & Social \\
\end{tabular}\end{center}
\end{table*}                    
  }

\subsection{Diagrams of the data and auxiliary tables}\label{sdia}
LOSD navigation may be assisted by diagrams which describe
the structure from each provenance.
Such diagrams are in the Supporting Information document
together with tables to facilitate the appreciation of the provided data.
A very simple example is given in Figure~\ref{dia} where the friendship
structure of the Facebook snapshots is revealed.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{../ontologies/facebook-legacy-AntonioAnzoategui18022013Friendship.ttl/draw}
    \caption{A diagram of the core structure that yields the friendship networks
    of the Facebook snapshots. A green edge denotes an OWL existential class restriction; an inverted nip denotes an OWL universal class restriction; a full (non-dashed) edge denotes an OWL functional property axiom (see Section~\ref{sont}). Further information and complete diagrams for each provenance are available in the Supporting Information document.}\label{dia}
\end{figure}

\subsection{SPARQL queries}\label{queries}
There are numerous useful and general purpose SPARQL queries to be performed against LOSD.
In this section, a selection of some of the most basic queries are made available due to their potential to be varied.
All queries assume the preamble \textttt{PREFIX po: <http://purl.org/socialparticipation/po/>}.
\begin{itemize}
  \item Retrieve the number of participants:
\begin{lstlisting}[language=spq]
SELECT (COUNT(DISTINCT ?author) as ?c) WHERE {
  ?author a po:Participant . 
}
\end{lstlisting}
  \item Retrieve the number of relations, be them interactions or
      friendships:
\begin{lstlisting}[language=spq]
SELECT (COUNT(?interaction) as ?c) WHERE {
  { ?interaction a po:Friendship } UNION { ?interaction a po:Interaction }
  UNION { ?interaction po:retweetOf ?message }
  UNION { ?interaction po:replyTo ?message }
  UNION { ?interaction po:directedTo ?participant }
}
\end{lstlisting}
            \item Retrieve all text produced by a specific user with URI \textttt{user\_uri}:
\begin{lstlisting}[language=spq]
SELECT (CONCAT(?text) as ?texts) WHERE {
  ?activity po:author <user_uri> . ?activity po:text ?text .
}
\end{lstlisting}
  \item List 1000 users (URIs and names) with the most friendships and the number of
      friendships in descending order by the number of friendships:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?participant (COUNT(?friendship) as ?c) WHERE {
  ?friendship a po:Friendship . ?friendship po:member ?participant . 
} ORDER BY DESC(?c) LIMIT 1000
\end{lstlisting}
  \item Retrieve text messages with the word ``crucial'' (case insensitive):
\begin{lstlisting}[language=spq]
SELECT ?text WHERE { 
  ?activity po:text ?text . FILTER regex(?text, 'crucial', 'i')
}
\end{lstlisting}
  \item List participants and respective full names whose name has the substring ``Amanda'':
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?participant ?name WHERE {
  ?participant po:observation ?obs . ?obs po:name ?name .
  FILTER regex(?name, 'Amanda', 'i') 
}
\end{lstlisting}
  \item Return all pairs of friends of a participant (with URI \textttt{participant\_uri}) which are friends themselves:
\begin{lstlisting}[language=spq]
SELECT DISTINCT ?friend1 ?friend2 WHERE {
  ?friendship1 po:member <participant_uri> . ?friendship1 po:member ?friend1 .
  ?friendship2 po:member <participant_uri> . ?friendship2 po:member ?friend2 .
  ?friendship3 po:member ?friend1 . ?friendship3 po:member ?friend2 .
}
\end{lstlisting}
  \item Return all interactions from replies in a snapshot:
\begin{lstlisting}[language=spq]
SELECT ?from ?to WHERE {
  ?message1 po:snapshot <snapshot_uri> .  ?message2 po:replyTo ?message1 .
  ?message1 po:author ?from .  ?message2 po:author ?to .
}
\end{lstlisting}
\end{itemize}

\subsubsection{Obtaining the networks}\label{snet}
A complex network $N=(n,l,m)$ may be regarded essentially as
a set of nodes $n$, a set of links $l$ that related the nodes,
and metadata $m$ such as node and link weights, text related to the nodes, etc.
This section presents queries for obtaining networks using LOSD
which are promptly envisioned by the authors.
Many other networks may be obtained because the criteria that entail
links, and restrains in the set of nodes and of links,
are determined by the research or application being developed.
Furthermore, we do not focus here on multilayer networks (e.g. bipartite networks), or on multigraphs (or muldimensional networks), i.e. on networks with more than one type of node or of link.
Here is a list with short descriptions of the networks and the corresponding SPARQL queries:
\begin{itemize}
  \item Overall facebook friendship network available in LOSD:
\begin{lstlisting}[language=spq]
